Reasonably Programmable Syntax

### 2017-05-28 popl18/syntax-popl18.tex
tweaks to intro

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
%Many languages define convenient list literal forms. This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language? 
When designing the surface syntax of a programming language, it is common practice to include literal forms for a select few standard data structures, e.g. lists. This practice neglects an unbounded assortment of other data structures for which literal forms might also be useful (consider, for example, sets, maps, matrices, regular expressions, HTML trees, URLs, SQL queries and chemical structures.) A more parsimonious approach would be to allow library providers to define new literal forms in a decentralized manner, but the mechanisms of  available to programmers today, e.g. \li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are, to put it pithily, \emph{unreasonable}. In particular,  they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.) Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, and therefore defeats much of the purpose of syntactic sugar. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.

This paper formally details a more reasonable alternative: \emph{typed literal macros} (TLMs). TLMs give library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types. Expansion is {strictly hygienic}, meaning 1) that expansions cannot make any assumptions about the application site context; and 2) that expansions do not reveal internal bindings to spliced terms or to the remainder of the program. 
% Partial parameter application lowers the syntactic cost of this strict style. 
This design ensures that clients are able to reason modularly about syntactic determinism and abstractly about the type and binding structure of the program. The system needs only to convey to clients (e.g. via secondary notation) the \emph{segmentation} of each literal, which gives the locations of the spliced segments. 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
% This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules.  This is the first detailed type-theoretic account of a typed, hygienic term-rewriting macro system for an ML-like language. % This calculus is the first such formalization of a typed macro system.

@@

% Put succinctly, we design a programming language in the ML tradition with a \emph{reasonably} programmable syntax.
\end{abstract}

@@

\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
When defining the surface syntax of a programming language, it is common practice to include a few ``literal forms'' specific to certain standard data structures.   
For example, Standard ML and other functional languages include sugar for constructing and pattern matching on lists, as defined in the standard library \cite{mthm97-for-dart}. For example, programmers can write \li{[x, y, z]} rather than \li{Cons(x, Cons(y, Cons(z, Nil)))}.

Third-party library providers are, quite justifiably, envious of the privileged status of these otherwise ordinary standard library definitions. After all, it is not difficult to find other examples of situationally useful library-specific syntactic sugar \cite{TSLs}. For example, 1) clients of a ``collections'' library might like syntactic sugar for finite sets and dictionaries; 2) clients of a ``web programming'' library might like syntactic sugar for HTML and JSON values; 3) a compiler writer might like syntactic sugar for the terms of the object language or various intermediate languages of interest; and 4) clients of a ``chemistry'' library might like syntactic sugar for chemical structures based on the SMILES standard \cite{anderson1987smiles}. This has motivated a long line of research into approaches that delegate more syntactic control to library providers.

\subsection{Syntax Dialects are Unreasonable}\label{sec:syntax-dialects-intro}
One well-worn approach is for library providers to use a syntax definition system to construct a library-specific \emph{syntax dialect}, i.e. a new syntax definition that extends the syntax definition given in the language definition with new derived forms (i.e. forms given meaning by expansion, rather than directly.) For example, Ur/Web defines a dialect of Ur's textual syntax with derived forms for SQL queries, XHTML elements and some other constructs defined in a  web programming library \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Figure \ref{fig:urweb} demonstrates how an XHTML expression containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2) can be expressed in Ur/Web. 
The desugarings of these derived forms, not shown, are substantially more verbose and, for programmers familiar with the standardized syntax for XHTML \cite{xhtml}, substantially more obscure.

\begin{figure}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-6px}
\caption{Derived XHTML forms with support for splicing at two different types are built in to Ur/Web.}
\vspace{-6px}
\label{fig:urweb}
\end{figure}
 % Such dialects are sometimes qualitatively taxonomized as amongst the ``domain-specific language'' for this reason \cite{fowler2010domain}. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
% Syntax definition systems , which we will discuss in Sec. \ref{sec:syntax-dialects}, have simplified the task of defining ``library-specific'' (a.k.a. ``domain-specific'') syntax dialects like Ur/Web, and have thereby contributed to their ongoing proliferation.

@@

\subsection{Contributions}
This work introduces a system of \textbf{typed literal macros (TLMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles. % comparable to the level of control they have when defining a syntax dialect.

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
%Many languages define convenient list literal forms. This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language? 
When designing the surface syntax of a programming language, it is common practice to build in literal forms for a select few standard data structures, e.g. lists. The problem is that this practice neglects an unbounded assortment of other data structures for which literal forms might also be useful (consider, for example, sets, maps, matrices, regular expressions, HTML trees, URLs, SQL queries and chemical structures.) A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. \li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.) Instead, they must reason transparently, i.e. about the underlying expansion. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.

This paper formally details a more reasonable alternative: \emph{typed literal macros} (TLMs). TLMs give library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types. Expansion is {strictly hygienic}, meaning 1) that expansions cannot make any assumptions about the application site context; and 2) that expansions do not reveal internal bindings to spliced terms or to the remainder of the program. 
% Partial parameter application lowers the syntactic cost of this strict style. 
This design ensures that clients are able to reason modularly about syntactic determinism and abstractly about the type and binding structure of the program. The system needs only to convey to clients (e.g. via secondary notation) the \emph{segmentation} of each literal, which gives the locations of the spliced segments. 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
% This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules.  This is the first detailed type-theoretic account of a typed, hygienic syntactic macro system for an ML-like language. % This calculus is the first such formalization of a typed macro system.

@@

% Put succinctly, we design a programming language in the ML tradition with a \emph{reasonably} programmable syntax.
\vspace{-10px}
\end{abstract}

@@

\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}

The surface syntax of a programming language serves as the user interface between human programmers and the language's semantics. As such, language designers often include various \emph{literal forms} in the surface syntax that are designed to decrease the cognitive cost of constructing and pattern matching over values of selected  types.

For example, in Standard ML (SML), the list expression literal \li{[x, y, z]} desugars to a term  semantically equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the expansion of literal forms is \emph{context independent}. We will return to the concept of context independence throughout this work.} List pattern literals are analagous. Another example is Ur/Web, which defines a dialect of Ur's textual syntax with expression and pattern literal forms for XHTML elements \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Figure \ref{fig:urweb} gives examles of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2). 
The desugarings of these literals, not shown, are substantially more verbose and, for programmers familiar with the standardized syntax for XHTML \cite{xhtml}, substantially more obscure.

Third-party library providers are quite justifiably envious of the privileged status of semantically ordinary types like those just discussed that have been equipped with convenient literal forms by a language designer. After all, it is not difficult to find other examples of types for which programmers have invented specialized syntax. For example, 1) clients of a ``collections'' library might like vector, matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library in SML might like Ur/Web style XHTML literals; 3) Ur/Web programmers might further like CSS literals, which Ur/Web does not provide; 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}. Research by \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use unstructured string encodings (a.k.a. ``stringly-typed programming'') and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}. \citet{TSLs} sampled string literals from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar (e.g. for SQL queries, regexes, file system paths, URLs, IP addresses and others.) Taken together, there appears to be a clear need for mechanisms that allow library providers to introduce new type-specific literal forms into a general-purpose language in a decentralized manner.

Unfortunately, the mechanisms available today are, to put it pithily, \emph{unreasonable}. The next two subsections flesh out this claim and thereby motivate the specific contribution of this paper, introduced in Sec. \ref{sec:contributions}: a decidedly more reasonable mechanism for defining new literal forms within an ML- or Scala-like general-purpose language (i.e. a language that has a rich static type system, support for pattern matching and an ML-style module system or Scala-style object system with support for type members.)

\subsection{Syntax Dialects are Unreasonable}\label{sec:syntax-dialects-intro}
One well-worn approach is for library providers to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new derived forms (i.e. forms given meaning by expansion, rather than directly.)

\begin{figure}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-6px}
\caption{XHTML literals with support for splicing at two different types are built in to Ur/Web.}
\vspace{-6px}
\label{fig:urweb}
\end{figure}
 % Such dialects are sometimes qualitatively taxonomized as amongst the ``domain-specific language'' for this reason \cite{fowler2010domain}. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
% Syntax definition systems , which we will discuss in Sec. \ref{sec:syntax-dialects}, have simplified the task of defining ``library-specific'' (a.k.a. ``domain-specific'') syntax dialects like Ur/Web, and have thereby contributed to their ongoing proliferation.

@@

\subsection{Contributions}\label{sec:contributions}
This work introduces a system of \textbf{typed literal macros (TLMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles. % comparable to the level of control they have when defining a syntax dialect.
